Estimating duration using survival analysis

### 2023-06-09 cis-perfect-testing.tex
Bit of editing and organising

Double interval censoring is a special case of censoring, a common and well-studied challenge encountered when dealing with survival data.
An event is \emph{censored} if the time of the event can only be bounded; it is \emph{left censored} if it is known to occur before a certain time, \emph{right censored} if it is known to occur after a certain time, and \emph{interval censored} if it is known to occur within an interval.
Interval censoring arises in CIS because we only observe the change from being not detectable (\ie: negative) at one test, to detectable (\ie: positive) at the following test and vice versa (see \cref{perf-test:fig:double-interval-censor}); \textcite{bogaertsSurvival} review many statistical methods applicable to interval censoring.
\emph{Double censoring} means that both the originating and terminating event are censored, and is also a well-studied problem, reviewed by \textcite{sunStatisticala}.
In CIS, we have \emph{double interval censoring} meaning that both the initiating and the terminating events are interval censored.
\todo{double interval censoring refs}.
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/double-interval-censor}
  \caption{Episodes data in CIS is double interval censored, meaning that both the start and end of the episode are only known up to an interval. Demonstrated here by a participant who is recorded negative at time 7 to positive at time 14, bounding the start of the episode; similarly, changing from positive at time 28 to negative at time 56 bounds the end of the episode. \label{perf-test:fig:double-interval-censor}}
\end{figure}

@@

Truncation is a large problem within CIS, especially when individuals move to the more infrequent testing schedule (described in \cref{E-intro:sec:cis}).
Consider that, based on the estimates in \cref{E-ATACCC}, the median length of an episode is 14 days \todo{check this number} we would not expect many episodes to last the 28 days that can occur between tests.
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/truncation}
  \caption{Episodes data in CIS is truncated, meaning that an unknown number of infections are never detected. Consider the two individuals shown here: we collect identical data for them (a series of negative tests) yet the top individual was never infected and the bottom individual was. See the main text for why this is an issue. \label{perf-test:fig:truncation}}
\end{figure}

@@

\begin{figure}
\includegraphics[width=\textwidth]{cis-perfect-testing/regions_diag}
\todo[inline]{Fix the impossible region to fill out the RHS; label the inadmissible regoin.}
\caption{Admissible, $\alpha_i$, inadmissible, $\beta_i$ and
untruncated, $\Omega_i^C$ regions for an episode $i$ in an
individual whose first test was at time 0 and negative, with subsequent
negative tests at times 7, 14, and 56, and positive tests at times 21
and 28. \label{perf-test:fig:partitionSpace}}
\end{figure}

@@

An uninformative prior would be $\lambda_t \dist \Beta(\alpha, \alpha)$, with common choices of $\alpha$ being $0.5$, Jeffreys' Prior, or $1$, a flat prior.
Flat or uninformative priors on the hazard lead to very informative priors on the survival, favouring short survival times (see \cref{perf-test:fig:flat-prior}).
The sharp falloff can be seen by noting that these priors have expectation 0.5, hence $\E(S(t)) = \E \left( \prod_{i=1}^{t-1} (1-\lambda_t) \right) = 0.5^{t-1}$ which decreases very quickly.
This prior expresses extreme scepticism in estimates from previous studies, such as \cref{E-ATACCC} and the meta-analysis of \textcite{cevikShedding}.
Therefore, I favour a prior which is weakly informative on the survival, or an informative prior used to represent our belief from the \cref{E-ATACCC} analysis.
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/flat-prior}
  \caption{The density of Beta(1, 1) flat prior and Beta(0.5, 0.5) uninformative priors (left) with the prior predictive survival time (right). The implied prior on the survival time is very short when using either prior on the hazard at each time. \label{perf-test:fig:flat-prior}}
\end{figure}

@@

A Beta(0.1, 1.9) prior has mean 0.05 while not having much information (the informativeness of a Beta distribution is related to $\alpha + \beta$ which here equals 2, the same as the flat prior case); and the central 95\% mass is 0.00--0.47.
The central estimate reflects previous estimates that the median duration is in the range 15--20 days~\autocite{cevikShedding}, which would translate to hazard of around 0.05.
The prior predictive distribution on the survival time implied by this prior is very vague (see \cref{perf-test:fig:vague-prior}).
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/vague-prior}
  \caption{The density of Beta(0.1, 1.9) vague prior (left) and the prior predictive survival time (right). The prior predictive survival time is very vague using a Beta(0.1, 1.9) prior. Note the different x-axis to \cref{perf-test:fig:flat-prior}. \label{perf-test:fig:vague-prior}}
\end{figure}

@@

There are two reasons to increase the uncertainty relative to the \cref{E-ATACCC} posterior.
First, we do not have full belief in these estimates due to the extrapolation problem previously discussed.
Second, the results may not fully generalise between the two studies.
The samples are tested using different laboratories, with different PCR machines, and hence may not have the same characteristics~\autocite{punchooLaboratory}.
The study participants, while both targeting a general population cohort, may also differ because ATACCC recruited through NHS Test and Trace while CIS recruited by sampling from a database of addresses (see \cref{E-intro:sec:data}).

@@

\begin{figure}
  \todo[inline]{Find figure comparing different duration distributions}
  \caption{Comparison of the duration distributions, see main text for details of how each is derived. \label{perf-test:fig:duration-dist}}
\end{figure}

@@

\begin{figure}
  \centering \includegraphics{cis-perfect-testing/survival-results}
  \caption{Comparison of the posterior estimate of the survival distribution (pointwise median and 95\% CrI) when using different priors on the hazard. \label{perf-test:fig:survival-results}}
\end{figure}

\begin{figure}
  \centering \includegraphics{cis-perfect-testing/hazard-results}
  \caption{Comparison of the posterior estimate of the hazard (pointwise median and 95\% CrI) when using different priors on the hazard. \label{perf-test:fig:hazard-results}}
\end{figure}

\begin{figure}
  \centering \includegraphics{cis-perfect-testing/hazard-pairs-results}
  \todo[inline]{Probably scrap this (correlation) plot}
  \caption{Correlation between posterior estimates of the hazard when using the vague prior. \label{perf-test:fig:hazard-pairs-results}}
\end{figure}

>>>>

Double interval censoring is a special case of censoring, a common and well-studied challenge encountered when dealing with survival data.
An event is \emph{censored} if the time of the event can only be bounded; it is \emph{left censored} if it is known to occur before a certain time, \emph{right censored} if it is known to occur after a certain time, and \emph{interval censored} if it is known to occur within an interval.
Interval censoring arises in CIS because we only observe the change from being not detectable (\ie: negative) at one test, to detectable (\ie: positive) at the following test and vice versa (see \cref{perf-test:fig:double-interval-censor}); \textcite{bogaertsSurvival} review many statistical methods applicable to interval censoring.
\emph{Double censoring} means that both the originating and terminating event are censored, and is also a well-studied problem, reviewed by \textcite{sunStatisticala}.
In CIS, we have \emph{double interval censoring} meaning that both the initiating and the terminating events are interval censored.
\todo{double interval censoring refs}.
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/double-interval-censor}
  \caption[Double-interval censoring in CIS data]{Episodes data in CIS is double interval censored, meaning that both the start and end of the episode are only known up to an interval. Demonstrated here by a participant who is recorded negative at time 7 to positive at time 14, bounding the start of the episode; similarly, changing from positive at time 28 to negative at time 56 bounds the end of the episode. \label{perf-test:fig:double-interval-censor}}
\end{figure}

@@

Truncation is a large problem within CIS, especially when individuals move to the more infrequent testing schedule (described in \cref{E-intro:sec:cis}).
Consider that, based on the estimates in \cref{E-ATACCC}, the median length of an episode is 14 days \todo{check this number} we would not expect many episodes to last the 28 days that can occur between tests.
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/truncation}
  \caption[Truncation in CIS data]{Episodes data in CIS is truncated, meaning that an unknown number of infections are never detected. Consider the two individuals shown here: we collect identical data for them (a series of negative tests) yet the top individual was never infected and the bottom individual was. See the main text for why this is an issue. \label{perf-test:fig:truncation}}
\end{figure}

@@

\begin{figure}
\includegraphics[width=\textwidth]{cis-perfect-testing/regions_diag}
\todo[inline]{Fix the impossible region to fill out the RHS; label the inadmissible regoin.}
\caption[Admissible, inadmissible, and untruncated infections]{Admissible, $\alpha_i$, inadmissible, $\beta_i$ and
untruncated, $\Omega_i^C$ regions for an episode $i$ in an
individual whose first test was at time 0 and negative, with subsequent
negative tests at times 7, 14, and 56, and positive tests at times 21
and 28. \label{perf-test:fig:partitionSpace}}
\end{figure}

@@

An uninformative prior would be $\lambda_t \dist \Beta(\alpha, \alpha)$, with common choices of $\alpha$ being $0.5$, Jeffreys' Prior, or $1$, a flat prior.
Flat or uninformative priors on the hazard lead to very informative priors on the survival, favouring short survival times (see \cref{perf-test:fig:flat-prior}).
The sharp falloff can be seen by noting that these priors have expectation 0.5, hence $\E(S(t)) = \E \left( \prod_{i=1}^{t-1} (1-\lambda_t) \right) = 0.5^{t-1}$ which decreases very quickly.
This prior expresses extreme scepticism in estimates from previous studies, such as \cref{E-ATACCC} and the meta-analysis of \textcite{cevikShedding}.
Therefore, I favour a prior which is weakly informative on the survival, or an informative prior used to represent our belief from the \cref{E-ATACCC} analysis.
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/flat-prior}
  \caption[Uninformative priors for the hazard]{The density of Beta(1, 1) flat prior and Beta(0.5, 0.5) uninformative priors (left) with the prior predictive survival time (right). The implied prior on the survival time is very short when using either prior on the hazard at each time. \label{perf-test:fig:flat-prior}}
\end{figure}

@@

A Beta(0.1, 1.9) prior has mean 0.05 while not having much information (the informativeness of a Beta distribution is related to $\alpha + \beta$ which here equals 2, the same as the flat prior case); and the central 95\% mass is 0.00--0.47.
The central estimate reflects previous estimates that the median duration is in the range 15--20 days~\autocite{cevikShedding}, which would translate to hazard of around 0.05.
The prior predictive distribution on the survival time implied by this prior is very vague (see \cref{perf-test:fig:vague-prior}).
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/vague-prior}
  \caption[Weakly informative priors for the hazard]{The density of Beta(0.1, 1.9) vague prior (left) and the prior predictive survival time (right). The prior predictive survival time is very vague using a Beta(0.1, 1.9) prior. Note the different x-axis to \cref{perf-test:fig:flat-prior}. \label{perf-test:fig:vague-prior}}
\end{figure}

@@

There are two reasons to increase the uncertainty relative to the \cref{E-ATACCC} posterior.
First, we do not have full belief in these estimates due to the extrapolation problem previously discussed.
Second, the results may not fully generalise between the two studies.
The samples are tested using different laboratories, with different PCR machines, and hence may not have the same characteristics~\autocite{punchooLaboratory}.
The study participants, while both targeting a general population cohort, may also differ because ATACCC recruited through NHS Test and Trace while CIS recruited by sampling from a database of addresses (see \cref{E-intro:sec:studies}).

@@

\begin{figure}
  \todo[inline]{Find figure comparing different duration distributions}
  \caption[Comparison of duration distributions]{Comparison of the duration distributions, see main text for details of how each is derived. \label{perf-test:fig:duration-dist}}
\end{figure}

@@

\begin{figure}
  \centering \includegraphics{cis-perfect-testing/survival-results}
  \caption[Comparison of survival function estimates under different priors]{Comparison of the posterior estimate of the survival distribution (pointwise median and 95\% CrI) when using different priors on the hazard. \label{perf-test:fig:survival-results}}
\end{figure}

\begin{figure}
  \centering \includegraphics{cis-perfect-testing/hazard-results}
  \caption[Comparison of hazard estimates under different priors]{Comparison of the posterior estimate of the hazard (pointwise median and 95\% CrI) when using different priors on the hazard. \label{perf-test:fig:hazard-results}}
\end{figure}

\begin{figure}
  \centering \includegraphics{cis-perfect-testing/hazard-pairs-results}
  \todo[inline]{Probably scrap this (correlation) plot}
  \caption[Correlation between hazard estimates]{Correlation between posterior estimates of the hazard when using the vague prior. \label{perf-test:fig:hazard-pairs-results}}
\end{figure}
